\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-23 manuscript/main.tex
model to the intro

	In modern biomedical studies, it is increasingly common to collect heterogeneous data modalities for each subject. As an illustrative example we use the Emory University renal study \citep{changBayesianLatentClass2020, jangPrincipalComponentAnalysis2021}. The data records both multiple renogram curves (functional data) and multiple renogram variables (vector data) per kidney, capturing distinct yet related physiological information. Such settings call for statistical methods that can jointly model functional and vector-valued covariates to account for both shared and modality-specific variation linked to clinical outcomes.

@@

	Without loss of generality, assume that both multivariate functional and scalar data are centered; that is, $E(X)=0$ and $E(\mathbf{Z})=\mathbf{0}$. Our goal is to predict a real outcome $Y$ based on $X$ and $\mathbf{Z}$ via the following functional regression model:

	\begin{equation}
		Y = \sum \limits_{r=1}^p \alpha_r Z_r + \sum \limits_{k=1}^K \int_0^1 \beta^{(k)}(t_k) X^{(k)}(t_k) dt_k + \epsilon = \langle \boldsymbol{\alpha}, \mathbf{Z} \rangle +  \langle \beta, X \rangle_{\mathcal{F}}  + \epsilon
		\label{eq: functional model}
	\end{equation}
	where $\boldsymbol{\alpha}=(\alpha_1, \ldots, \alpha_p)^\top \in \mathbb{R}^p$ and $\beta=(\beta^{(1)}, \ldots, \beta^{(K)}) \in \mathcal{F}$ are respectively the regression coefficient vector and function that characterize the effect of scalar and functional predictors on the outcome.

@@

Instead of solely maximizing the squared empirical covariance
 $\widehat{\textnormal{Cov}}^2 
\bigl(
\langle \widetilde{W}, \xi \rangle_\mathcal{H}, Y 
\bigr)
$, 
we incorporate this roughness penalty to simultaneously control the complexity of the estimated functional components.

One possible approach is to extend the smoothed functional PCA framework of \citet{Rice1991} by modifying the objective in \eqref{def:maximizer_squared_empirical_cov} to
	\begin{equation*}
	\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)
	- \sum \limits_{j=1}^K \lambda_j 
	\operatorname{PEN}({\xi}_j)
	,
\end{equation*}
where the smoothing parameters $\{\lambda_j\}_{j=1}^K$ control the trade-off between maximizing covariance and penalizing roughness.

\medskip
\noindent
\textit{Proposed estimation procedure.}~
However, the above approach of \citet{Rice1991} assumes that the functional predictors admit an orthogonal expansion in the $L^2$ sense. As an alternative, we adopt the strategy of \citet{Silverman1996}, which replaces the standard orthonormality constraint with a weaker one based on a modified inner product that incorporates roughness. Accordingly, our estimation procedure at the $l$-th iteration, iteration index omitted and assuming the observations have been residualized in previous steps, solves the following optimization problem:
	\begin{equation}\label{def:maximizer_squared_empirical_cov_reg}
	\hat{\xi} 
	= 
	(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}
	)
	:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
	}~\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
	\bigr)~s.t.~\| \xi  \|_\mathcal{H} +  \sum \limits_{j=1}^K \lambda_j 
	\operatorname{PEN}(\xi_K) = 1.
\end{equation}
This formulation maximizes the squared covariance over a class of smooth functions.
Proposition~\ref{proposition:eigen_regul} expresses this problem as a generalized Rayleigh quotient.

>>>>

Modern biomedical studies frequently collect diverse data types from each subject.
As an illustrative example, the Emory University renal study \citep{changBayesianLatentClass2020, jangPrincipalComponentAnalysis2021}  records both multiple renogram curves (functional data) and multiple renogram variables (vector data)  for each kidney. These modalities provide distinct yet related physiological signals. To effectively analyze such heterogeneous data, we need statistical methods capable of jointly modeling functional and vector-valued covariates. This approach accounts for both shared and modality-specific variations linked to clinical outcomes. Our work focuses on developing a method to fit the following hybrid functional linear regression model:
\begin{equation}	\label{eq: functional model}
	Y_i
	= 
	\boldsymbol{\beta}^\top \mathbf{Z}_i
	+ 
	\sum 
	\limits_{j=1}^K 
	\int  \beta_j(t) X_{ij}(t) dt + \epsilon_i,
	\quad
	i = 1, \ldots, n,
	%= \langle \boldsymbol{\alpha}, \mathbf{Z} \rangle +  \langle \beta, X \rangle_{\mathcal{F}}  + \epsilon
\end{equation}
where $Y$ is a scalar response, $\mathbf{Z}$ is the Euclidean vector covariate and 
$X_{i1}(t), \ldots, X_{iK}(t)$ are functional predicters.
For notational convenience, we assume throguth this paper that the responsd sand preedictoras ahve been centered so the intercept can be ignored. 
%where $\boldsymbol{\alpha}=(\alpha_1, \ldots, \alpha_p)^\top \in \mathbb{R}^p$ and $\beta=(\beta^{(1)}, \ldots, \beta^{(K)}) \in \mathcal{F}$ are respectively the regression coefficient vector and function that characterize the effect of scalar and functional predictors on the outcome.

@@

@@

	Instead of solely maximizing the squared empirical covariance
	$\widehat{\textnormal{Cov}}^2 
	\bigl(
	\langle \widetilde{W}, \xi \rangle_\mathcal{H}, Y 
	\bigr)
	$, 
	we incorporate this roughness penalty to simultaneously control the complexity of the estimated functional components.
	One possible approach is to extend the smoothed functional PCA framework of \citet{Rice1991} by modifying the objective in \eqref{def:maximizer_squared_empirical_cov} to
	\begin{equation*}
		\widehat{\textnormal{Cov}}^2 
		\bigl(
		\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
		\bigr)
		- \sum \limits_{j=1}^K \lambda_j 
		\operatorname{PEN}({\xi}_j)
		,
	\end{equation*}
	where the smoothing parameters $\{\lambda_j\}_{j=1}^K$ control the trade-off between maximizing covariance and penalizing roughness. 
	However, this approach  \citet{Rice1991} assumes that the functional predictors admit an orthogonal expansion in the $L^2$ sense.

	\medskip
	\noindent
	\textit{Proposed estimation procedure.}~
	To avoid the orthogonal basis assumption of  \citet{Rice1991}, we adopt the strategy of \citet{Silverman1996}, which replaces the standard orthonormality constraint with a weaker one based on a modified inner product that incorporates roughness. Accordingly, our estimation procedure at the $l$-th iteration, iteration index omitted and assuming the observations have been residualized in previous steps, solves the following optimization problem:
	\begin{equation}\label{def:maximizer_squared_empirical_cov_reg}
		\hat{\xi} 
		= 
		(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}
		)
		:= \arg \max_{\xi \in \widetilde{\mathcal{H}}
		}~\widehat{\textnormal{Cov}}^2 
		\bigl(
		\langle \widetilde{W}, \xi \rangle_{\mathcal{H}}, Y 
		\bigr)~s.t.~\| \xi  \|_\mathcal{H} +  \sum \limits_{j=1}^K \lambda_j 
		\operatorname{PEN}(\xi_K) = 1.
	\end{equation}
	This formulation maximizes the squared covariance over a class of smooth functions.
	Proposition~\ref{proposition:eigen_regul} expresses this problem as a generalized Rayleigh quotient.
